Safe Transferable Regions

### 2016-07-01 paper/introduction.tex
introduction

The motivating example in Fig.~\ref{fig:motivating-eg} represents
a \C{SELECT} query operator for streaming data. 
The operator receives a stream of input messages, each
associated with a time (window) $t$, processed by method
\texttt{onReceive}.  Each input message contains a list of inputs,
each of which is processed by applying a user-defined function to
create a corresponding output.  The operator may receive multiple
messages with the same timestamp (and messages with different
timestamps may be delivered out of order). A timing-message (an
invocation of method \texttt{OnNotify}) indicates that no more input
messages with a timestamp $t$ will be subsequently delivered.  At this
point, the operator completes the processing for time window $t$ and
sends a corresponding output message to its successor.

@@

This example (of  a producer creating a data structure and transferring it to a consumer)
is an instance of a general idiom of a set of concurrent actors interacting via messages
(representing data-structures).
Our focus in this paper is a region-based approach to realizing such transferable
data, motivated by the following concerns.

@@

% Region-based memory management, both manual as well as automatic, has
% been known for a long time. Manual region-based memory management
% suffers from the usual drawbacks, namely the potential for invalid
% references and the consequent lack of memory safety. Automatic
% region-based memory management systems guarantee memory safety, but
% impose various restrictions.  MLKit, which implements the approach
Safe region-based memory management using types was
pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, who 
use only lexically scoped regions.  At runtime, the set of all regions
(existing at a point in time) forms a stack. Thus, the lifetimes of
all regions must be well-nested: it is not possible to have two
regions whose lifetimes overlap, with neither one's lifetime contained
within the other.  Unfortunately, the data structures in the above
example do not satisfy this restriction (as the output messages for
multiple time windows may be simultaneously live, without any
containment relation between their lifetimes).  We refer to regions
with lexically scoped lifetimes as \emph{stack regions} and to regions
that do not have such a lexically scoped region as \emph{dynamic}
regions.
%
Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
that can be safely transferred across address spaces. We refer to such dynamic
regions as \emph{transferable} regions.

Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
that combine linear types and regions to suport dynamic regions.
Unlike the prior work, we ensure memory safety in the presence of
transferable regions through a combination of a static
typing discipline and lightweight runtime checks (in place of linear types).
The cornerstone of this approach is an \C{open} lexical block for transferable regions,
that ``opens'' a transferable region and guarantees that the region
won't be transferred/freed while it is open. By nesting an
MLKit-style \C{letregion} lexical block, that delimits
the lifetime of a stack region, inside an \C{open} lexical block for a
transferable region, we can guarantee that the transferable region
will remain live as long as the stack region is live. We say that the
former \emph{outlives} the later., and any references from the stack
region to the transferable region are therefore safe.  Next, we note
that by controlling the outlives relationships between various
regions, we only allow safe cross-region references, while prohibiting
unsafe ones. In the above example, an outlives relationship
\emph{from} the stack region \emph{to} the transferable region means
that the references in that direction are allowed, but not the
references in the opposite direction. In contrast, if an \C{open}
block of a transferable region $R_0$ is nested inside an \C{open}
block of another transferable region $R_1$, we do not establish any
outlives relationships, thus declaring our intention to not allow any
cross-region references between $R_0$ and $R_1$.  Finally, we observe
that outlives relationships are established based on the lexical
structure of the program, hence a static type system can enforce them
effectively. By assigning region types to objects, which capture the
regions such objects are allocated in, and by maintaining outlives
relationships between various regions, we can statically decide the
safety of all references in the program.

Ensuring memory safety using this approach requires ensuring
that the use of transferable regions satisfies certain temporal properties.
Firstly, a transferable region cannot be transferred/freed
inside an open block of that region (i.e, while it is still open).
Secondly, a transferred/freed region cannot be opened. These are
typestate invariants on the transferable region objects, which are
hard to enforce statically in the presence of unrestricted
aliasing. Techniques like linear types and unique pointers can be used
to restrict aliasing, but the constraints they impose are often hard
to program around. We therefore enforce typestate invariants at
runtime via lightweight checks. In particular, we define an acceptable
state transition discipline for transferable regions
(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
transition of a transferable region (\eg, from \emph{open} state to
\emph{freed} state) is valid or not. The check is lightweight since it
only involves checking a single tag that captures the current state.
We believe that this is a reasonable choice since regions are
coarse-grained objects manipulated infrequently, when compared to the
fine-grained objects that are present inside these regions, for which
safety is enforced statically.  An added advantage of delegating the
enforcement of typestate invariants to runtime is that our region type
system is simple, which made it possible to formulate a type inference
that completely eliminates the need to write region type annotations.
This, we believe, significantly reduces the impediment to adopt our
approach in practical setting.

>>>>

In this paper, we present a \emph{safe} and \emph{efficient} region-based approach
to \emph{transferring data} between a set of concurrent actors, motivated by the following concerns.

Consider the example (of  a producer creating a data structure and transferring it to a consumer)
in Fig.~\ref{fig:motivating-eg}. This represents a \C{SELECT} query operator for streaming data. 
The operator receives a stream of input messages, each
associated with a time (window) $t$, processed by method
\texttt{onReceive}.  Each input message contains a list of inputs,
each of which is processed by applying a user-defined function to
create a corresponding output.  The operator may receive multiple
messages with the same timestamp (and messages with different
timestamps may be delivered out of order). A timing-message (an
invocation of method \texttt{OnNotify}) indicates that no more input
messages with a timestamp $t$ will be subsequently delivered.  At this
point, the operator completes the processing for time window $t$ and
sends a corresponding output message to its successor.

@@

@@

% Region-based memory management, both manual as well as automatic, has
% been known for a long time. Manual region-based memory management
% suffers from the usual drawbacks, namely the potential for invalid
% references and the consequent lack of memory safety. Automatic
% region-based memory management systems guarantee memory safety, but
% impose various restrictions.  MLKit, which implements the approach
Safe region-based memory management using types was
pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, who 
use only lexically scoped regions.  At runtime, the set of all regions
(existing at a point in time) forms a stack. Thus, the lifetimes of
all regions must be well-nested: it is not possible to have two
regions whose lifetimes overlap, with neither one's lifetime contained
within the other.  Unfortunately, the data structures in the above
example do not satisfy this restriction (as the output messages for
multiple time windows may be simultaneously live, without any
containment relation between their lifetimes).  We refer to regions
with lexically scoped lifetimes as \emph{stack regions} and to regions
that do not have such a lexically scoped region as \emph{dynamic}
regions.

Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
that can be safely transferred across address spaces. We refer to such dynamic
regions as \emph{transferable} regions.
%
Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
that combine linear types and regions to suport dynamic regions.
Unlike the prior work, we ensure memory safety in the presence of
transferable regions through a \emph{combination of a static
typing discipline and lightweight runtime checks in place of linear types}.

The cornerstone of this approach is an \C{open} lexical block for transferable regions,
that ``opens'' a transferable region and guarantees that the region
won't be transferred/freed while it is open. By nesting an
MLKit-style \C{letregion} lexical block, that delimits
the lifetime of a stack region, inside an \C{open} lexical block for a
transferable region, we can guarantee that the transferable region
will remain live as long as the stack region is live. We say that the
former \emph{outlives} the later., and any references from the stack
region to the transferable region are therefore safe.  Next, we note
that by controlling the outlives relationships between various
regions, we only allow safe cross-region references, while prohibiting
unsafe ones. In the above example, an outlives relationship
\emph{from} the stack region \emph{to} the transferable region means
that the references in that direction are allowed, but not the
references in the opposite direction. In contrast, if an \C{open}
block of a transferable region $R_0$ is nested inside an \C{open}
block of another transferable region $R_1$, we do not establish any
outlives relationships, thus declaring our intention to not allow any
cross-region references between $R_0$ and $R_1$.  Finally, we observe
that outlives relationships are established based on the lexical
structure of the program, hence a static type system can enforce them
effectively. By assigning region types to objects, which capture the
regions such objects are allocated in, and by maintaining outlives
relationships between various regions, we can statically decide the
safety of all references in the program.

Ensuring memory safety using this approach requires ensuring
that the use of transferable regions satisfies certain temporal properties.
Firstly, a transferable region should not be transferred/freed
inside an open block of that region (i.e, while it is still open).
Secondly, a transferred/freed region should not be opened. These are
typestate invariants on the transferable region objects, which are
hard to enforce statically in the presence of unrestricted
aliasing. Techniques like linear types and unique pointers can be used
to restrict aliasing, but the constraints they impose are often hard
to program around. We therefore enforce typestate invariants at
runtime via lightweight checks. In particular, we define an acceptable
state transition discipline for transferable regions
(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
transition of a transferable region (\eg, from \emph{open} state to
\emph{freed} state) is valid or not. The check is lightweight since it
only involves checking a single tag that captures the current state.
We believe that this is a reasonable choice since regions are
coarse-grained objects manipulated infrequently, when compared to the
fine-grained objects that are present inside these regions, for which
safety is enforced statically.

One of the novel contributions of this paper is a type inference algorithm
that eliminates the need for users to write region type annotations.
This, we believe, significantly reduces the impediment to adopt our
approach in practical setting.
